Minimal trap spaces of Boolean networks are maximal siphons of their Petri net encoding

### 2022-12-23 manuscripts/TCS2022/tcs2022.tex
Computation of fixed points

\end{definition}
A \emph{marking} for a Petri net is a mapping \(m : P \mapsto \mathbb{N}\) that assigns a number of tokens to each place. 
A place \(p\) is marked by a marking \(m\) if and only if \(m(p) > 0\). 
Marking \(m\) can be seen as a subset of \(P\) that contains all marked places by \(m\).
We shall write \(pred(x)\) (resp.\ \(succ(x)\)) to represent the set of vertices that have a (non-zero weighted) arc leading to (resp.\ coming from) \(x\).
In this work, we consider a class of Petri nets called 1-safe Petri nets where every place has at most 1 token and all arcs are of weight 1.
In this case, weights are implicitly omitted in the arcs of a Petri net.
Then, a transition \(t \in T\) is \emph{enabled} at a marking \(m\) if and only if \(pred(t) \subseteq m\). 
The firing of \(t\) leads to a new marking \(m'\) specified by \(m' = (m \backslash pred(t)) \cup succ(t)\).
Note that when multiple transitions are enabled, we need to embed one firing scheme (similar to the update scheme of a Boolean network) to the Petri net.
The classical firing scheme is that only one of the enabled transition is non-deterministically chosen to fire~\cite{Murata1989}.

@@

Second, fixed points // TODO:

@@

\begin{table}[!htb]
  \caption{Timing comparisons between bioLQM (LQM), PyBoolNet (PBN), mpbn and the four variants of Trappist on selected models from the literature.}
  \centering%
  \label{tab:result_real}
  \begin{tabular}{rlrrrrrrrrr}
    \toprule
    & & & & & & & \multicolumn{4}{c}{Trappist}\\
    \cmidrule(rr){8-11}
    & model & \(n\) & \(|M|\) & LQM & PBN & mpbn & SAT & CLP & ILP & ASP \\
    \midrule % n < 100
    \rownb & Arabidopsis\_thaliana~\cite{DesignPrinciplesGeneNetworks} & 15 & 8 & 0.15 & - & NM & - & - & - & - \\
    \rownb & Bcell~\cite{Dutta2019} & 73 & 72 & 0.21 & - & - & - & - & - & - \\
    \rownb & Cacace\_TdevModel~\cite{Cacace2020} & 61 & 28 & 1.21 & - & NM & - & - & - & - \\
    \rownb & Corral\_ThIL17diff~\cite{corral2021interplay} & 92 & $>1000$ & N/A & - & - & - & - & - & - \\
    \rownb & DNA\_damage~\cite{DesignPrinciplesGeneNetworks} & 26 & 16 & 0.23 & - & NM & - & - & - & - \\
    \rownb & Drosophila~\cite{RodrguezVega2014} & 52 & 128 & 0.33 & - & - & - & - & - & - \\
    \rownb & EMT~\cite{Rozum2021} & 69 & 268 & - & - & - & - & - & - & - \\
    \rownb & mast\_cell~\cite{aghamiri2020automated} & 73 & $>1000$ & N/A & - & - & - & - & - & - \\
    \rownb & FT-GRN~\cite{ChvezHernndez2022} & 23 & 32 & DNF & - & NM & - & - & - & - \\
    \rownb & hedgehog~\cite{DesignPrinciplesGeneNetworks} & 65 & $>1000$ & N/A & - & - & - & - & - & - \\
    \rownb & metastatic~\cite{DesignPrinciplesGeneNetworks} & 10 & 4 & 0.11 & - & NM & - & - & - & - \\
    \rownb & Pancreatic\_Cancer~\cite{DesignPrinciplesGeneNetworks} & 43 & $>1000$ & N/A & - & - & - & - & - & - \\
    \rownb & Pluripotent~\cite{DesignPrinciplesGeneNetworks} & 36 & 276 & 0.37 & - & NM & - & - & - & - \\
    \rownb & Rho-GTPases~\cite{DesignPrinciplesGeneNetworks} & 33 & 2 & 0.21 & - & - & - & - & - & - \\
    \rownb & Pluripotency~\cite{YachieKinoshita2018} & 36 & 440 & DNF & - & NM & - & - & - & - \\
    \rownb & p53\_high\_dna~\cite{DesignPrinciplesGeneNetworks} & 16 & 1 & 0.40 & - & NM & - & - & - & - \\
    \rownb & p53\_low\_dna~\cite{DesignPrinciplesGeneNetworks} & 16 & 1 & 0.38 & - & NM & - & - & - & - \\

>>>>

\end{definition}
A \emph{marking} for a Petri net is a mapping \(m : P \mapsto \mathbb{N}\) that assigns a number of tokens to each place. 
A place \(p\) is marked by a marking \(m\) if and only if \(m(p) > 0\). 
Marking \(m\) can be seen as a subset of \(P\) that contains all marked places by \(m\).
We shall write \(pred(x)\) (resp.\ \(succ(x)\)) to represent the set of vertices that have a (non-zero weighted) arc leading to (resp.\ coming from) \(x\).
In this work, we consider a class of Petri nets called 1-safe Petri nets where every place has at most 1 token and all arcs are of weight 1.
In this case, weights are implicitly omitted in the arcs of a Petri net.
Then, a transition \(t \in T\) is \emph{enabled} at a marking \(m\) if and only if \(pred(t) \subseteq m\). 
A marking \(m\) is called a deadlock if there are no enabled transitions at \(m\).
The firing of \(t\) leads to a new marking \(m'\) specified by \(m' = (m \backslash pred(t)) \cup succ(t)\).
Note that when multiple transitions are enabled, we need to embed one firing scheme (similar to the update scheme of a Boolean network) to the Petri net.
The classical firing scheme is that only one of the enabled transition is non-deterministically chosen to fire~\cite{Murata1989}.

@@

Second, we consider fixed points in Boolean networks.
Let \(s\) be a fixed point of a Boolean network \(\mathcal{N}\).
We have a subspace \(m\) corresponding to \(s\) as follows: \(\forall v \in V, m(v) = s(v)\), i.e., all nodes are fixed in \(m\).
Clearly, \(s\) is a trap set of \(\mathcal{N}\) regardless of the update scheme.
Hence, \(m\) is a trap space of \(\mathcal{N}\).
In addition, since \(\left\vert S_{\mathcal{N}}[m]\right\vert = 1\), \(m\) is also a minimal trap space.
To compute all fixed points of \(\mathcal{N}\), we can add more constraints to the encoded ASP characterizing all conflict-free siphons (equivalently trap spaces).
For every \(v \in V\), we add to the encoded ASP the rule
\[
  \verb|p-v| ; \verb|n-v|.
\]
that ensures that for every conflict-free siphon \(S\), it contains either \(\verb|p-v|\) or \(\verb|n-v|\) for every \(v \in V\).
Equivalently, the trap space corresponding to \(S\) is always a fixed point.
Now, the set of answer sets of the encoded ASP is equivalent to the set of fixed points of \(\mathcal{N}\).
In particular, when solving the encoded ASP using an ASP solver, we do not need to use the built-in option for computing set-inclusion maximal answer sets.
Note that we can also build another ASP characterizing all fixed points of \(\mathcal{N}\) based on the equivalence between a fixed point of \(\mathcal{N}\) and a deadlock of its Petri net encoding~\cite{chatain2014characterization}.
This approach may give a more compact ASP.

@@

\begin{table}[!htb]
  \caption{Timing comparisons between bioLQM (LQM), PyBoolNet (PBN), mpbn and the four variants of Trappist on selected models from the literature.}
  \centering%
  \label{tab:result_real}
  \begin{tabular}{rlrrrrrrrrr}
    \toprule
    & & & & & & & \multicolumn{4}{c}{Trappist}\\
    \cmidrule(rr){8-11}
    & model & \(n\) & \(|M|\) & LQM & PBN & mpbn & SAT & CLP & ILP & ASP \\
    \midrule % n < 100
    \rownb & Arabidopsis\_thaliana~\cite{DesignPrinciplesGeneNetworks} & 15 & 8 & 0.15 & - & NM & - & - & - & - \\
    \rownb & Bcell~\cite{Dutta2019} & 73 & 72 & 0.21 & - & - & - & - & - & - \\
    \rownb & Cacace\_TdevModel~\cite{Cacace2020} & 61 & 28 & 1.21 & - & NM & - & - & - & - \\
    \rownb & Corral\_ThIL17diff~\cite{corral2021interplay} & 92 & $>1000$ & N/A & - & - & - & - & - & - \\
    \rownb & DNA\_damage~\cite{DesignPrinciplesGeneNetworks} & 26 & 16 & 0.23 & - & NM & - & - & - & - \\
    \rownb & Drosophila~\cite{RodrguezVega2014} & 52 & 128 & 0.33 & - & - & - & - & - & - \\
    \rownb & EMT~\cite{Rozum2021} & 69 & 268 & 39.22 & - & - & - & - & - & - \\
    \rownb & mast\_cell~\cite{aghamiri2020automated} & 73 & $>1000$ & N/A & - & - & - & - & - & - \\
    \rownb & FT-GRN~\cite{ChvezHernndez2022} & 23 & 32 & DNF & - & NM & - & - & - & - \\
    \rownb & hedgehog~\cite{DesignPrinciplesGeneNetworks} & 65 & $>1000$ & N/A & - & - & - & - & - & - \\
    \rownb & metastatic~\cite{DesignPrinciplesGeneNetworks} & 10 & 4 & 0.11 & - & NM & - & - & - & - \\
    \rownb & Pancreatic\_Cancer~\cite{DesignPrinciplesGeneNetworks} & 43 & $>1000$ & N/A & - & - & - & - & - & - \\
    \rownb & Pluripotent~\cite{DesignPrinciplesGeneNetworks} & 36 & 276 & 0.37 & - & NM & - & - & - & - \\
    \rownb & Rho-GTPases~\cite{DesignPrinciplesGeneNetworks} & 33 & 2 & 0.21 & - & - & - & - & - & - \\
    \rownb & Pluripotency~\cite{YachieKinoshita2018} & 36 & 440 & DNF & - & NM & - & - & - & - \\
    \rownb & p53\_high\_dna~\cite{DesignPrinciplesGeneNetworks} & 16 & 1 & 0.40 & - & NM & - & - & - & - \\
    \rownb & p53\_low\_dna~\cite{DesignPrinciplesGeneNetworks} & 16 & 1 & 0.38 & - & NM & - & - & - & - \\
